\bf Hierarchical Network Structure Promotes Dynamical Robustness

### 2015-07-23 tex/discussion.tex
minor edits

%!TEX root = ../paper.tex
% Our development suggests two predictions that could be used to test the theory described here. One is that,
\section{Conclusion}
Our analysis predicts that, in general, an ensemble of systems where robustness has been the predominant object of selection and has been positively selected over a sufficiently long period of time should exhibit a bias toward more hierarchical network topologies. Given the manner in which we define robustness in \ref{eq:robustnessfromstoch}, this is a very general constraint. In the short term, this prediction may be further evaluated at the levels of both metabolic and transcription factor networks, which have already been shown to display hierarchical structure, but whose dynamics have not been sufficiently well characterized to ascertain their dynamical robustness as we have defined it here \cite{Zhao2006,Bhardwaj2010,Colm,Prill2005}. Indeed, Figure 2 from \cite{Prill2005} demonstrates an enrichment of various hierarchical network architectures consisting of multiple SCCs relative to those containing cycles with single SCCs for networks with three and four variables in \emph{E. coli}, \emph{S. cerevisiae}, and \emph{D. melanogaster} transcription factor networks as well as the signal transduction knowledge environment network and the \emph{C. elegans} neural network. The network architecture that we predict to be enriched by selection for robustness for three variable networks ranks 4th out of 13 in overall abundance across these networks and within each network individually with the exception of the \emph{C. elegans} neural network, and the three more abundant network architectures are all proper subgraphs of this one. These results are essentially consistent with our prediction, though, of course, we would not claim that there could not be various alternative explanations for these observations. At the ecological level, a system subjected to the environmental stress of overfishing, which may imply selection for robustness, has been observed to exhibit such a bias toward more hierarchical network architectures \cite{Bascompte2005}.
%The other prediction of the theory outlined here that may be able to be checked is the converse statement that an ensemble of systems demonstrating a positive or negative bias in their network topology toward hierarchy may have been under selection for, or respectively against, robustness. Of course, in this case, factors other than selection for robustness would not be ruled out by this analysis alone.
In the long term, this prediction may be evaluated using experimental evolution by comparing the degree of hierarchy that emerges in the evolution of gene regulatory network topology in the context of both static and fluctuating environments that impose differential selection strengths for dynamical robustness \cite{Leroi1994}.

>>>>

%!TEX root = ../paper.tex
% Our development suggests two predictions that could be used to test the theory described here. One is that,
\section{Conclusion}
Our analysis predicts that, in general, an ensemble of systems where robustness has been the predominant object of selection and has been positively selected over a sufficiently long period of time should exhibit a bias toward more hierarchical network topologies. Given the manner in which we define robustness in \ref{eq:robustnessfromstoch}, this is a very general constraint. In the short term, this prediction may be further evaluated at the levels of both metabolic and transcription factor networks, which have already been shown to display hierarchical structure, but whose dynamics have not been sufficiently well characterized to ascertain their dynamical robustness as we have defined it here \cite{Zhao2006,Bhardwaj2010,Colm,Prill2005}.

Indeed, Figure 2 from \cite{Prill2005} demonstrates an enrichment of various hierarchical network architectures consisting of multiple SCCs relative to those containing cycles with single SCCs for networks with three and four variables in \emph{E. coli}, \emph{S. cerevisiae}, and \emph{D. melanogaster} transcription factor networks as well as the signal transduction knowledge environment network and the \emph{C. elegans} neural network. The network architecture that we predict to be most enriched by selection for robustness for three variable networks ranks fourth out of thirteen in overall abundance across these empirically observed classes of networks and within each class of network individually. In each case, the three most abundant network architectures are all proper subgraphs of the fourth-ranked most hierarchical one. These observations are consistent with both our analytical and numerical predictions, though, of course, we do not claim that there could not be alternative explanations for the relative enrichment of more hierarchical network architectures in these systems.

At the ecological level, a system subjected to the environmental stress of overfishing, which may imply selection for robustness, has been observed to exhibit such a bias toward more hierarchical network architectures \cite{Bascompte2005}.
%The other prediction of the theory outlined here that may be able to be checked is the converse statement that an ensemble of systems demonstrating a positive or negative bias in their network topology toward hierarchy may have been under selection for, or respectively against, robustness. Of course, in this case, factors other than selection for robustness would not be ruled out by this analysis alone.
In the long term, our prediction may be evaluated using experimental evolution by comparing the degree of hierarchy that emerges in the evolution of gene regulatory network topology in the context of both static and fluctuating environments that impose differential selection strengths for dynamical robustness \cite{Leroi1994,Hindre2012}.

### 2015-07-23 tex/figures.tex
minor edits

\begin{figure*}[!ht]
\centering
\noindent\includegraphics[width=1.0\textwidth]{fig/modsccsym.pdf}
\caption{{\bf Open systems, strongly connected components and symmetries of robustness.} (A) Example of the combination of open system modules to construct closed systems. (B) SCCs highlighted in gray for each of the four graphs representing the interdependencies relevant to four different three variable systems. The most hierarchical network, top panel, is the one that maximizes the number of SCCs and the number of links between them.
% We therefore define hierarchy as $max(\hbox{ED}) - \hbox{ED}$ where ED is the edit distance representing the number of link addition/deletion operations necessary to transform a given graph into the most hierarchical one.
The two panels in the middle represent examples of hierarchical modular systems that posess both modularity (i.e. SCCs with more than one variable) and hierarchy. (C) Symmetries of the $\hier$ transformation between graphs and SCCs. The transformation $a$ represents an interchange of SCCs, $b$ moving a link between nodes in a component and $c$ adding a link. All three transformations represent symmetries of the $\hier$ transformation from graphs to SCCs while only $a$ and $b$ are symmetries of robustness.}
\label{fig:modsccsym}
\end{figure*}

>>>>

\begin{figure*}[!ht]
\centering
\noindent\includegraphics[width=1.0\textwidth]{fig/modsccsym.pdf}
\caption{{\bf Open systems, strongly connected components and symmetries of robustness.} (A) Example of the combination of open system modules to construct closed systems. (B) SCCs highlighted in gray for each of the four graphs representing the interdependencies relevant to four different three variable systems. The most hierarchical network, top panel, maximizes the number of links between SCCs.
% We therefore define hierarchy as $max(\hbox{ED}) - \hbox{ED}$ where ED is the edit distance representing the number of link addition/deletion operations necessary to transform a given graph into the most hierarchical one.
The two panels in the middle represent examples of hierarchical modular systems that posess both modularity (i.e. SCCs with more than one variable) and hierarchy. (C) Symmetries of the $\hier$ transformation that maps a graph representing interactions among system variables to a graph representing the relationships among the SCCs of the system. The transformation $a$ represents an interchange of SCCs, $b$ moving a link between nodes in a component and $c$ adding a link. All three transformations represent symmetries of the $\hier$ transformation from graphs to SCCs while only $a$ and $b$ are symmetries that preserve dynamical robustness.}
\label{fig:modsccsym}
\end{figure*}

### 2015-07-23 tex/overview.tex
minor edits

\subsection{Most hierarchical network architectures maximize dynamical robustness}
Following the examples of the previous section, here we show, in general, that dynamical robustness is maximized for a given collection of SCCs by maximizing the number of links between them. This is accomplished by demonstrating that any graph $G$ with fewer than the maximum, $l_{\mathrm{max}}$, of links between the given collection of SCCs must have a lower dynamical robustness than the graph, $G_{\mathrm{max}}$, with $l=l_{\mathrm{max}}$ and that this $G_{\mathrm{max}}$ is indeed the graph $G_{\mathrm{tot}}$ which gives the total ordering on the SCCs. Examining the expression in \ref{eq:robschematic} and noting that $R_{\alpha}$ are all strictly less than one proves that networks maximizing $l$, will also maximize $R$.  Given two connected components $C_{\alpha}$ and $C_{\beta}$ with $v_{\alpha}$ and $v_{\beta}$ nodes respectively, we have a maximum of $v_{\alpha} v_{\beta}$ links going from $C_{\alpha}$ to $C_{\beta}$.  Hence, $l \le \sum_{(\alpha,\beta) \in \hier(G)} v_{\alpha} v_{\beta}$.  Since every acyclic digraph can be embedded into a totally ordered

@@

This argument also works when relationships between multiple system variables are perturbed simultaneously,

@@

resample $m$ interactions.  Then the analogue of \ref{eq:singleresamplemutau} is

@@

m_0, m_1, \ldots. m_n}$ ways of choosing $m_i$ links from $C_i$ and

@@

This analytical result implies that the interaction graphs for systems that are the most robust will maximize the number of links between a given set of SCCs. Because this result is purely topological in nature, it does not depend at all upon any particular details such as the probability distribution from which the variable interaction strengths are sampled or the size of the system. Any network whose associated dynamical system has the interaction graph among its SCCs equivalent to the total ordering is more robust than those associated to any of the other interaction graphs in \reffigscc{}. The graph associated to the total ordering is the most hierarchical network architecture for any given number of SCCs like that of \reffigscc{} top for three variable systems where the \emph{highest} variable in the hierarchy has directed links to all other variables in the network, the second highest variable has directed links to all other variables in the network except the highest one, \emph{et cetera} (\refsupp{}). The result that dynamical robustness is correlated with network hierarchy therefore applies to an even broader class of dynamical systems than the particular random ensembles we have studied directly.

\subsection{Examples of the analytical relationship between hierarchy and robustness}

To test the prediction of the analytical results in \ref{eq:robschematic} and \ref{eq:robustnessmultiple}, we computed approximations to the probability distribution of stability and dynamical robustness relative to network architecture for ensembles of systems having two or three interacting variables (see \refsupp{} \ref{tab:structstabmat} and \ref{tab:structstabmat3}). For all of these, we found that robustness is correlated with connectivity, but that the most robust systems have intermediate connectivity for a given network size (\reffigrobustconnect). Accounting for the number of cycles in a network architecture reveals a strong correlation between robustness and connectivity that was hidden when networks with any number of cycles were considered together (\reffigconnectcycle3D3x3). While the most hierarchical network architecture will always lack cycles altogether, cycle number alone is clearly insufficient to account for robustness as the members of each class span nearly the entire range of possible robustness values. Consistent with our analysis of the symmetries of robustness, we found that the most hierarchical network architecture is the most robust (\reffigrobusthierarchy). Moreover, if we consider hierarchy partitioned by connectivity, we find that there is a monotonic increase in robustness following any line of increasing hierarchy in \reffigconnectdist3D3x3.

>>>>

\subsection{Most hierarchical network architectures maximize dynamical robustness}
Following the examples of the previous section, here we show, in general, that dynamical robustness is maximized for a given collection of SCCs by maximizing the number of links between them. This is accomplished by demonstrating that any graph $G$ with $l < l_{\mathrm{max}}$ ($l_{\mathrm{max}}$ refers to the maximum possible number of links between the given collection of SCCs) must have a lower dynamical robustness than one of the graphs $G_{\mathrm{max}}$ having $l=l_{\mathrm{max}}$, and that these $G_{\mathrm{max}}$ have SCCs related by the graph $G_{\mathrm{tot}}$ which gives the total ordering on the given set of SCCs.

Examining the expression in \ref{eq:robschematic} and noting that $R_{\alpha}$ are all strictly less than one proves that networks maximizing $l$ for a given set of SCCs, will also maximize $R$.  Given two connected components $C_{\alpha}$ and $C_{\beta}$ with $v_{\alpha}$ and $v_{\beta}$ nodes respectively, we have a maximum of $v_{\alpha} v_{\beta}$ links going from $C_{\alpha}$ to $C_{\beta}$.  Hence, $l \le \sum_{(\alpha,\beta) \in \hier(G)} v_{\alpha} v_{\beta}$.  Since every acyclic digraph can be embedded into a totally ordered

@@

The results of the previous sections also hold when relationships between multiple system variables are perturbed simultaneously,

@@

resample $m$ interactions, as a model for cases in which multiple interaction strengths among system variables are modified in a single timestep of the evolutionary process.  Then the analog of \ref{eq:singleresamplemutau} is

@@

m_0, m_1, \ldots. m_n}$ ways of choosing $m_i$ links from within $C_i$ and

@@

This analytical result implies that the interaction graphs for systems that are the most robust will maximize the number of links between a given set of SCCs. Because this result is purely topological in nature, it does not depend at all upon any particular details such as the size of the system or the probability distribution from which the variable interaction strengths are sampled. Any network whose associated dynamical system has the interaction graph among its SCCs equivalent to the total ordering is more robust than those associated to any of the other interaction graphs in \reffigscc{}. The graph associated to the total ordering is the most hierarchical network architecture for any given number of SCCs like that of \reffigscc{} top for three variable systems where the \emph{highest} variable in the hierarchy has directed links to all other variables in the network, the second highest variable has directed links to all other variables in the network except the highest one, \emph{et cetera} (\refsupp{} \ref{sec:totalordering}). The result that dynamical robustness is correlated with network hierarchy therefore applies to an even broader class of dynamical systems than the particular random ensembles we have studied directly.

\subsection{Examples of the relationship between hierarchy and robustness}

To test the prediction of the analytical results in \ref{eq:robschematic} and \ref{eq:robustnessmultiple}, we computed approximations to the probability distribution of stability and dynamical robustness relative to network architecture for ensembles of systems having two or three interacting variables (see \refsupp{} \ref{tab:structstabmat} and \ref{tab:structstabmat3}). For all of these, we found that robustness is correlated with connectivity, but that the most robust systems have intermediate connectivity for a given network size (\reffigrobustconnect). Accounting for the number of cycles in a network architecture reveals a strong correlation between robustness and connectivity that was hidden when networks with any number of cycles were considered together (\reffigconnectcycle3D3x3). While the most hierarchical network architecture will always lack cycles altogether, cycle number alone is clearly insufficient to account for robustness as the members of each class span nearly the entire range of possible robustness values. Consistent with our analysis, we found that the most hierarchical network architecture is the most robust (\reffigrobusthierarchy). Moreover, if we consider hierarchy partitioned by connectivity, we find that there is a monotonic increase in robustness following any line of increasing hierarchy in \reffigconnectdist3D3x3.
